Future challenges for Research Software Engineering

### 2025-04-16 future_challenges_for_RSEng.tex
work in feedback from PR

Looking ahead over a 10 year timeframe, we hope that the RSE community will have
supported a number of changes and developments that help to make research software outputs
more sustainable, robust and maintainable. We see this happening in a number of ways. However,
we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.
With development of software tools or scripts becoming an increasingly important part of everyday
research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
do see the work of RSEs developing and shifting as the research community changes.

The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
We would, therefore, not expect domain researchers to also become experts in a range of
advanced RSE topics. 
At the same time, we do recognise that the application of core software development best practices
to enhance and address problems with research codebases, that can often make up the bulk of
current RSE work, will change. As noted above, these are skills that we
would expect any researcher who writes code to have in 10 years time. 
The rapid development of AI means that researchers are likely to have access to tooling that can assist
them in writing robust, sustainable code, lowering the barrier to developing code, or at least changing the
required skill set to some extent.
They can also be expected to have an understanding of frameworks to support testing, packaging and
deployment of their code. Again, all these skills will, we anticipate, be gained as part of training
infrastructure that will provide domain scientists with software development expertise as part of their
core scientific training.

So where do RSEs fit into this picture? We expect to see RSEs providing technical skills and input
to research activities in range of different areas, including the following:
\begin{itemize}
  \item High-quality, robust architecture and design for research software
  \item Development of software for novel and emerging architectures
  \item Application of specialist numerical and statistical methods within software
  \item Optimisation of codes for use on high-performance computing infrastructure
  \item Green computing - efficient implementations of algorithms/methods and use of hardware
  \item ...
\end{itemize}

\subsection{Definition and Evolution of RSEs}
\begin{framed}
How will the definition of an RSE evolve in ten years?

What will not change in the essence of an RSE ten years from now?
\end{framed}
% Here we consider how the topics of RSEng will change in the face of the digitalization of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.

>>>>

Looking ahead over a 10 year timeframe, we hope that the RSE community will have
supported a number of changes and developments that help to make research software outputs
more sustainable, robust and maintainable. We see this happening in a number of ways. However,
we want to ensure that one of the core achievements made by the RSE community in this time
frame is putting the infrastructure and training opportunities in place, alongside the necessary
advocacy, to ensure that researchers receive core software skills as part of their scientific training.

With development of software tools or scripts becoming an increasingly important part of everyday
research activities, this is vital and it will allow RSEs to focus on other more advanced and specialist
areas. In this context, we do not foresee the demise of RSE as an independent role. However, we
do see the work of RSEs developing and shifting as the research community changes.

The specialist skills that RSEs provide take a significant amount of time to develop and maintain.
We would, therefore, not expect domain researchers to also become experts in a range of
advanced RSE topics. 
At the same time, we do recognise that the application of core software development best practices
to enhance and address problems with research codebases, that can often make up the bulk of
current RSE work, will change. As noted above, these are skills that we
would expect any researcher who writes code to have in 10 years time. 
The rapid development of AI means that researchers are likely to have access to tooling that can assist
them in writing robust, sustainable code, lowering the barrier to developing code, or at least changing the
required skill set to some extent.
They can also be expected to have an understanding of frameworks to support testing, packaging and
deployment of their code as well as integration with central domain specific infrastructures.
Again, all these skills will, we anticipate, be gained as part of training infrastructure that will
provide domain scientists with software development expertise as part of their core scientific training.

So where do RSEs fit into this picture? We expect to see RSEs providing technical skills and input
to research activities in a range of different areas, including the following:
\begin{itemize}
  \item High-quality, robust architecture and design for research software
  \item Seamless integration with research and data infrastructures
  \item Development of software for novel and emerging architectures
  \item Application of specialist numerical and statistical methods within software
  \item Optimisation of codes for use on high-performance computing infrastructure
  \item Green computing - efficient implementations of algorithms/methods and use of hardware
  \item ...
\end{itemize}

\subsection{Definition and Evolution of RSEs}
\begin{framed}
\hfill How will the definition of an RSE evolve in ten years?\\

\hfill What will not change in the essence of an RSE ten years from now?
\end{framed}
% Here we consider how the topics of RSEng will change in the face of the digitalization of society.
% Also here: What will our topics be after version control doesn't need to be taught anymore?
% Digitalisation will also feature more heavily in the domain curricula.
